---
title: "Summative Assessment 1"
author: "Karl Isaiah Buenafe"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  warning    = FALSE,
  message    = FALSE,
  fig.align  = "center",
  fig.width  = 10,
  fig.height = 6,
  cache      = FALSE
)
```


```{r load-libraries, message=FALSE, warning=FALSE, paged.print=FALSE, eval=TRUE , echo=FALSE}
library(tidyverse)    
library(caret)       
library(rpart)      
library(rpart.plot)     
library(randomForest) 
library(glmnet)      
library(corrplot)      
library(gridExtra)      
library(scales)      
library(knitr)         
library(kableExtra)
library(FactoMineR)   
library(factoextra)    
library(ROCR)           
library(e1071)  

theme_set(theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle = element_text(colour = "grey50", hjust = 0.5),
    panel.grid.minor = element_blank()
  ))

set.seed(42)
```

---

# Unit 1: R for Data Mining {.tabset}

## 1.1 Introduction to Modern Data Mining

### Load & Inspect the Dataset

```{r load-data}
df = read.csv("customer_churn.csv", stringsAsFactors = FALSE)

cat("Dataset Dimensions:", nrow(df), "rows ×", ncol(df), "columns\n")

head(df,10)
```

```{r structure}
str(df)
```

```{r missing-values}

missing_summary = data.frame(
  Variable     = names(df),
  Type         = sapply(df, class),
  Missing      = sapply(df, function(x) sum(is.na(x))),
  Missing_Pct  = round(sapply(df, function(x) mean(is.na(x))) * 100, 2),
  Unique_Values = sapply(df, function(x) length(unique(x)))
)

kable(missing_summary, row.names = FALSE,
      caption = "Table 1: Variable dictionary with missing value audit") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  column_spec(3, color = ifelse(missing_summary$Missing > 0, "red", "green"))
```

```{r descriptive-stats}
numeric_vars = df %>% select(Tenure, MonthlyCharges, TotalCharges)

desc_stats = numeric_vars %>%
  summarise(across(everything(), list(
    n      = ~n(),
    mean   = ~round(mean(.), 2),
    sd     = ~round(sd(.), 2),
    min    = ~round(min(.), 2),
    q25    = ~round(quantile(., 0.25), 2),
    median = ~round(median(.), 2),
    q75    = ~round(quantile(., 0.75), 2),
    max    = ~round(max(.), 2)
  ))) %>%
  pivot_longer(everything(), names_to = c("Variable", "Stat"), names_sep = "_(?=[^_]+$)") %>%
  pivot_wider(names_from = Variable, values_from = value)

kable(desc_stats, caption = "Table 2: Descriptive statistics for continuous features") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r churn-distribution}
churn_dist <- df %>%
  count(Churn) %>%
  mutate(Percentage = round(n / sum(n) * 100, 1))

kable(churn_dist, col.names = c("Churn", "Count", "Percentage (%)"),
      caption = "Table 3: Churn target variable distribution") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

The data set contains **10,000 telecom subscriber records** with numeric and non-numeric variables. Conducting manual analysis on large data set prohibits us from extracting accurate insights that could be found within thousands of rows. However, if we conduct data mining, we can achieve the following:

- **Discover hidden patterns**: Non-obvious interactions between tenure, charges, and contract type
- **Build predictive scores**: Probabilistic churn risk models for retention targeting
- **Quantify feature importance**: Identify which variables matter most via ensemble methods
- **Handle collinearity**: TotalCharges is highly correlated with Tenure (r ≈ 0.83), requiring regularised methods


---

## 1.2 Data Visualization

### Churn Rate by Categorical Variables

```{r viz-categorical, fig.height=9}

contract_churn <- df %>%
  group_by(Contract, Churn) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(Contract) %>%
  mutate(pct = round(n / sum(n) * 100, 1))

internet_churn <- df %>%
  group_by(InternetService, Churn) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(InternetService) %>%
  mutate(pct = round(n / sum(n) * 100, 1))

senior_churn <- df %>%
  mutate(SeniorLabel = ifelse(SeniorCitizen == 1, "Senior", "Non-Senior")) %>%
  group_by(SeniorLabel, Churn) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(SeniorLabel) %>%
  mutate(pct = round(n / sum(n) * 100, 1))

churn_colors <- c("No" = "#2196F3", "Yes" = "#F44336")

p1 <- ggplot(contract_churn, aes(x = Contract, y = pct, fill = Churn)) +
  geom_col(position = "dodge", width = 0.65) +
  geom_text(aes(label = paste0(pct, "%")), position = position_dodge(0.65),
            vjust = -0.4, size = 3.5, fontface = "bold") +
  scale_fill_manual(values = churn_colors) +
  scale_y_continuous(labels = label_percent(scale = 1), limits = c(0, 90)) +
  labs(title = "Churn by Contract Type", x = NULL, y = "Percentage (%)") +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))

p2 <- ggplot(internet_churn, aes(x = InternetService, y = pct, fill = Churn)) +
  geom_col(position = "dodge", width = 0.65) +
  geom_text(aes(label = paste0(pct, "%")), position = position_dodge(0.65),
            vjust = -0.4, size = 3.5, fontface = "bold") +
  scale_fill_manual(values = churn_colors) +
  scale_y_continuous(labels = label_percent(scale = 1), limits = c(0, 90)) +
  labs(title = "Churn by Internet Service", x = NULL, y = "Percentage (%)")

p3 <- ggplot(senior_churn, aes(x = SeniorLabel, y = pct, fill = Churn)) +
  geom_col(position = "dodge", width = 0.55) +
  geom_text(aes(label = paste0(pct, "%")), position = position_dodge(0.55),
            vjust = -0.4, size = 3.5, fontface = "bold") +
  scale_fill_manual(values = churn_colors) +
  scale_y_continuous(labels = label_percent(scale = 1), limits = c(0, 90)) +
  labs(title = "Churn by Senior Status", x = NULL, y = "Percentage (%)")

grid.arrange(p1, p2, p3, ncol = 3,
             top = "Figure 1: Churn Rates Across Key Categorical Variables")
```

> **Insight**: The chart shows a significant difference of churning rate across each categorical value. It shows a lower but uniformed churning rate (~25%-28%). This suggest the validity of conducting multivariate analysis.

---

### Distribution of Numeric Features by Churn

```{r viz-distributions, fig.height=8, message=FALSE, warning=FALSE}
p4 <- ggplot(df, aes(x = Tenure, fill = Churn, colour = Churn)) +
  geom_density(alpha = 0.45, size = 0.8) +
  scale_fill_manual(values = churn_colors) +
  scale_colour_manual(values = churn_colors) +
  labs(title = "Tenure Distribution by Churn",
       x = "Tenure (Months)", y = "Density")

p5 <- ggplot(df, aes(x = MonthlyCharges, fill = Churn, colour = Churn)) +
  geom_density(alpha = 0.45, size = 0.8) +
  scale_fill_manual(values = churn_colors) +
  scale_colour_manual(values = churn_colors) +
  labs(title = "Monthly Charges Distribution by Churn",
       x = "Monthly Charges ($)", y = "Density")

grid.arrange(p4, p5, ncol = 2,
             top = "Figure 2: Numeric Feature Distributions by Churn Status")
```

> **Insight**: Across continuous variable, it shows the similarity of the churning rate.

---

### Correlation Matrix

```{r viz-correlation, fig.height=5, fig.width=6}
df_temp <- df %>%
  mutate(ChurnBinary = as.integer(Churn == "Yes"))

corr_data <- df_temp %>%
  select(Tenure, MonthlyCharges, TotalCharges, ChurnBinary)

corr_matrix <- cor(corr_data)

corrplot(corr_matrix,
         method     = "color",
         type       = "upper",
         addCoef.col = "black",
         number.cex  = 0.9,
         tl.col      = "black",
         tl.srt      = 45,
         col         = colorRampPalette(c("#2196F3", "white", "#F44336"))(200),
         title       = "Figure 3: Correlation Matrix",
         mar         = c(0, 0, 2, 0))
```

> **Insight**: The figure above shows a **strong positive correlation** between Tenure and Total Charges (r = 0.77) and a **moderate positive correlation** between Monthly Charge and Total Charge. These values suggest that Tenure has a stronger influence on the total charge of customers. 

---

### Churn Rate by Tenure Bucket

```{r viz-tenure-bucket, fig.height=5}
df_temp <- df %>%
  mutate(
    TenureBucket = cut(Tenure,
                       breaks = c(0, 12, 24, 36, 48, 60, 71),
                       labels = c("0-12", "13-24", "25-36", "37-48", "49-60", "61-71"),
                       include.lowest = TRUE),
    ChurnBinary = as.integer(Churn == "Yes")
  )

tenure_churn <- df_temp %>%
  group_by(TenureBucket) %>%
  summarise(ChurnRate = round(mean(ChurnBinary) * 100, 1), .groups = "drop")

ggplot(tenure_churn, aes(x = TenureBucket, y = ChurnRate, fill = ChurnRate)) +
  geom_col(width = 0.65, colour = "white") +
  geom_text(aes(label = paste0(ChurnRate, "%")), vjust = -0.5,
            fontface = "bold", size = 4) +
  scale_fill_gradient(low = "#90CAF9", high = "#F44336", guide = "none") +
  scale_y_continuous(limits = c(0, 35)) +
  labs(title    = "Figure 4: Churn Rate by Tenure Bucket",
       subtitle = "Near-uniform churn across all tenure groups",
       x        = "Tenure Bucket (Months)",
       y        = "Churn Rate (%)")
```
> **Insight**: The figure shows a uniformity of churning rate across Tenure buckets. However it shows that the highest churning rate is observed within the 61-71 Tenure bucket. The figure above suggest that there is no correlation between Tenure and the Churning Rate. 

---

## 1.3 Data Transformation

```{r data-transformation}
cat("Total missing values:", sum(is.na(df)), "\n")

df_clean <- df %>%
  mutate(
    Gender         = as.factor(Gender),
    Partner        = as.factor(Partner),
    Dependents     = as.factor(Dependents),
    PhoneService   = as.factor(PhoneService),
    InternetService = as.factor(InternetService),
    Contract       = as.factor(Contract),
    Churn          = as.factor(Churn)
  )

cat("Factor variables:\n")
cat(paste(names(df_clean)[sapply(df_clean, is.factor)], collapse = ", "), "\n")

df_clean <- df_clean %>%
  mutate(
    Tenure_scaled         = as.numeric(scale(Tenure)),
    MonthlyCharges_scaled = as.numeric(scale(MonthlyCharges)),
    TotalCharges_scaled   = as.numeric(scale(TotalCharges))
  )

scaling_check <- df_clean %>%
  summarise(
    Tenure_mean   = round(mean(Tenure_scaled), 4),
    Tenure_sd     = round(sd(Tenure_scaled), 4),
    Charges_mean  = round(mean(MonthlyCharges_scaled), 4),
    Charges_sd    = round(sd(MonthlyCharges_scaled), 4)
  )

scaling_check
```

---

## 1.4 Data Wrangling

```{r data-wrangling}

Q1  <- quantile(df_clean$TotalCharges, 0.25)
Q3  <- quantile(df_clean$TotalCharges, 0.75)
IQR <- Q3 - Q1

lower_fence <- Q1 - 1.5 * IQR
upper_fence <- Q3 + 1.5 * IQR

df_clean <- df_clean %>%
  filter(TotalCharges >= lower_fence & TotalCharges <= upper_fence)

cat(sprintf("Rows retained: %d (removed %d outliers, %.2f%%)\n",
            nrow(df_clean),
            10000 - nrow(df_clean),
            (10000 - nrow(df_clean)) / 10000 * 100))


df_clean <- df_clean %>%
  mutate(
    ChargesPerMonth = TotalCharges / (Tenure + 1),
    LongTermContract = as.integer(Contract != "Month-to-month"),
    ChurnBinary = as.integer(Churn == "Yes")
  )

cat("New derived variables: ChargesPerMonth, LongTermContract, ChurnBinary\n")

```

```{r wrangling-summary}
cat("Final dataset dimensions:", nrow(df_clean), "rows ×", ncol(df_clean), "columns\n\n")

churn_final <- df_clean %>%
  count(Churn) %>%
  mutate(Pct = round(n / sum(n) * 100, 1))

churn_final
```

---

# Unit 2: Tuning Predictive Models {.tabset}

## 2.1 Model Complexity

### Feature Matrix Preparation

```{r feature-matrix}
df_model <- df_clean %>%
  mutate(
    Gender_enc          = as.integer(Gender == "Male"),
    Partner_enc         = as.integer(Partner == "Yes"),
    Dependents_enc      = as.integer(Dependents == "Yes"),
    PhoneService_enc    = as.integer(PhoneService == "Yes"),
    InternetService_enc = case_when(
      InternetService == "No"         ~ 0L,
      InternetService == "DSL"        ~ 1L,
      InternetService == "Fiber optic" ~ 2L
    ),
    Contract_enc = case_when(
      Contract == "Month-to-month" ~ 0L,
      Contract == "One year"       ~ 1L,
      Contract == "Two year"       ~ 2L
    )
  )

feature_cols <- c("Tenure", "MonthlyCharges", "TotalCharges", "SeniorCitizen",
                  "Gender_enc", "Partner_enc", "Dependents_enc",
                  "PhoneService_enc", "InternetService_enc", "Contract_enc")

X <- df_model[, feature_cols]
y <- df_model$Churn 

preProc  <- preProcess(X, method = c("center", "scale"))
X_scaled <- predict(preProc, X)

cat("Feature matrix:", nrow(X_scaled), "×", ncol(X_scaled), "\n")
cat("Target distribution:", table(y), "\n")
```

### Decision Tree

```{r decision-tree}
dt_model <- rpart(
  Churn ~ Tenure + MonthlyCharges + TotalCharges + SeniorCitizen +
          Gender_enc + Partner_enc + Dependents_enc +
          PhoneService_enc + InternetService_enc + Contract_enc,
  data    = df_model,
  method  = "class",
  control = rpart.control(maxdepth = 5, cp = 0.001)
)

rpart.plot(dt_model,
           type    = 4,
           extra   = 104,
           fallen.leaves = TRUE,
           main    = "Figure 5: Decision Tree (max depth = 5)",
           cex     = 0.7,
           tweak   = 1.1)
```

### Logistic Regression

```{r logistic-regression-basic}

lr_model <- glm(
  Churn ~ Tenure + MonthlyCharges + TotalCharges + SeniorCitizen +
          Gender_enc + Partner_enc + Dependents_enc +
          PhoneService_enc + InternetService_enc + Contract_enc,
  data   = df_model,
  family = binomial(link = "logit")
)

cat("Logistic Regression summary (coefficient count):", length(coef(lr_model)), "\n")
cat("AIC:", round(AIC(lr_model), 2), "\n")
```

### Complexity Comparison Table

```{r complexity-table}
complexity_df <- data.frame(
  Attribute = c("Hypothesis Space", "Parameters", "Handles Non-linearity",
                "Handles Collinearity", "Interpretability", "Regularisation",
                "Overfitting Risk", "Best Use Case"),
  DecisionTree = c(
    "Axis-aligned recursive splits",
    "Up to 2^depth − 1 split rules",
    "Yes — captures interactions",
    "Tolerant — selects best split",
    "High — visual tree",
    "Pruning (max_depth, cp, min_split)",
    "High at large depth",
    "Non-linear, interaction-heavy data"
  ),
  LogisticRegression = c(
    "Linear decision boundary",
    "p + 1 coefficients (11 here)",
    "No — strictly linear",
    "Sensitive — inflated SE under collinearity",
    "High — coefficient signs & magnitudes",
    "L1 (Lasso) or L2 (Ridge) penalty",
    "Low — underfits non-linear patterns",
    "Linearly separable, interpretable output needed"
  )
)

kable(complexity_df,
      col.names = c("Attribute", "Decision Tree (depth=5)", "Logistic Regression"),
      caption   = "Table 8: Model Complexity Comparison") %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  column_spec(1, bold = TRUE)
```

---

## 2.2 Bias–Variance Trade-Off

### Conceptual Framework

The total expected generalisation error decomposes as:

$$\text{Total Error} = \text{Bias}^2 + \text{Variance} + \text{Irreducible Noise}$$

```{r bias-variance-illustration, fig.height=4}

complexity <- seq(1, 10, length.out = 100)
bias2      <- 10 * exp(-0.5 * complexity)
variance   <- 0.2 * exp(0.4 * complexity)
total      <- bias2 + variance + 2  

bv_df <- data.frame(
  Complexity = rep(complexity, 3),
  Error      = c(bias2, variance, total),
  Component  = rep(c("Bias²", "Variance", "Total Error"), each = 100)
)

ggplot(bv_df, aes(x = Complexity, y = Error, colour = Component, linetype = Component)) +
  geom_line(size = 1.2) +
  scale_colour_manual(values = c("Bias²" = "#F44336", "Variance" = "#2196F3",
                                  "Total Error" = "#4CAF50")) +
  annotate("vline", xintercept = 3.5, linetype = "dashed", colour = "grey40") +
  annotate("text", x = 3.7, y = 9, label = "Optimal\nComplexity",
           colour = "grey30", size = 3.5) +
  labs(title    = "Figure 6: Bias–Variance Trade-Off",
       subtitle = "Total error is minimised at the optimal model complexity",
       x        = "Model Complexity →",
       y        = "Expected Error",
       colour   = "Component",
       linetype = "Component") +
  theme(legend.position = "right")
```



> **In this dataset**: All models achieve ~70–73% accuracy because a **majority-class classifier** (always predict "No") achieves 72.9% accuracy. 
---

## 2.3 Cross-Validation (k = 10)

```{r cross-validation}
ctrl <- trainControl(
  method          = "cv",
  number          = 10,
  classProbs      = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final"
)


set.seed(42)
dt_cv <- train(
  x          = X_scaled,
  y          = y,
  method     = "rpart",
  trControl  = ctrl,
  metric     = "ROC",
  tuneLength = 5
)


set.seed(42)
lr_cv <- train(
  x         = X_scaled,
  y         = y,
  method    = "glm",
  family    = "binomial",
  trControl = ctrl,
  metric    = "ROC"
)

cat("Decision Tree best cp:", dt_cv$bestTune$cp, "\n")
cat("Logistic Regression AUC:", round(max(lr_cv$results$ROC), 4), "\n")
```

```{r cv-metrics}

get_cv_metrics <- function(model_obj, model_name) {
  preds <- model_obj$pred
  preds <- preds[preds$cp == model_obj$bestTune$cp | is.null(model_obj$bestTune$cp), ]
  cm    <- confusionMatrix(preds$pred, preds$obs, positive = "Yes")
  data.frame(
    Model     = model_name,
    Accuracy  = round(cm$overall["Accuracy"] * 100, 2),
    Precision = round(cm$byClass["Precision"] * 100, 2),
    Recall    = round(cm$byClass["Recall"] * 100, 2),
    F1_Score  = round(cm$byClass["F1"] * 100, 2),
    AUC       = round(max(model_obj$results$ROC), 4)
  )
}

dt_metrics <- get_cv_metrics(dt_cv, "Decision Tree (depth=5)")
lr_metrics <- get_cv_metrics(lr_cv, "Logistic Regression")

dt_metrics
lr_metrics
```

> **Interpretation**: Both models achieve ~72–73% accuracy but near-zero F1 for the churn class The AUC values above 0.5 indicate some discriminative ability, but **Precision, Recall, and F1 are the correct evaluation criteria** for an imbalanced churn prediction task.

---

## 2.4 Classification — Random Forest with Grid Search

```{r random-forest, cache=TRUE}

rf_grid <- expand.grid(mtry = c(2, 3, 5, 7))

ctrl_rf <- trainControl(
  method          = "cv",
  number          = 10,
  classProbs      = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final"
)

set.seed(42)
rf_cv <- train(
  x          = X_scaled,
  y          = y,
  method     = "rf",
  trControl  = ctrl_rf,
  tuneGrid   = rf_grid,
  metric     = "ROC",
  ntree      = 100
)

cat("Best mtry:", rf_cv$bestTune$mtry, "\n")
print(rf_cv$results[, c("mtry", "ROC", "Sens", "Spec")])
```

```{r rf-feature-importance, fig.height=5}

importance_df <- varImp(rf_cv)$importance %>%
  rownames_to_column("Feature") %>%
  arrange(desc(Overall)) %>%
  mutate(Feature = reorder(Feature, Overall))

ggplot(importance_df, aes(x = Feature, y = Overall, fill = Overall)) +
  geom_col(width = 0.7) +
  coord_flip() +
  scale_fill_gradient(low = "#90CAF9", high = "#1E5799", guide = "none") +
  geom_text(aes(label = round(Overall, 1)), hjust = -0.2, size = 3.5) +
  scale_y_continuous(limits = c(0, max(importance_df$Overall) * 1.15)) +
  labs(title    = "Figure 7: Random Forest Feature Importances",
       x        = NULL,
       y        = "Importance Score")
```

```{r rf-final-metrics}

rf_metrics <- get_cv_metrics(rf_cv, "Random Forest (tuned)")
rf_metrics
```

> **Final Model**: The **Random Forest** is the only model that detects churned customers meaningfully, despite lower accuracy (69.3%). it's able to catch at-risk customers which justifies the accuracy trade-off. MonthlyCharges (31%), TotalCharges (30%), and Tenure (21%) account for 82% of total importance, with service and demographic features contributing the remaining 18%.

---

# Unit 3: Regression-Based Methods {.tabset}

## 3.1 Logistic Regression

```{r logit-three-vars}
logit_3 <- glm(
  ChurnBinary ~ Tenure + MonthlyCharges + TotalCharges,
  data   = df_model,
  family = binomial(link = "logit")
)

summary(logit_3)
```

```{r logit-table}
logit_coefs <- as.data.frame(summary(logit_3)$coefficients)
logit_coefs$CI_Low  <- confint(logit_3)[, 1]
logit_coefs$CI_High <- confint(logit_3)[, 2]
logit_coefs$OddsRatio <- exp(coef(logit_3))
logit_coefs <- round(logit_coefs, 4)
names(logit_coefs) <- c("Coefficient", "Std Error", "z-Value", "p-Value", "CI Low", "CI High", "Odds Ratio")

kable(logit_coefs,
      caption = "Table 12: Logistic Regression Results — Tenure, MonthlyCharges, TotalCharges") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  row_spec(1, background = "#EEF4FB") %>%
  column_spec(4, bold = TRUE,
              color = ifelse(logit_coefs[["p-Value"]] < 0.05, "#C62828", "black"))
```

**Coefficient Interpretation:**

| Variable | β | Odds Ratio | Interpretation |
|---|---|---|---|
| Intercept | −1.055 | 0.348 | Baseline churn log-odds when all predictors = 0 |
| Tenure | +0.003 | 1.003 | Each additional month → negligible change in churn odds (*p* = 0.37) |
| MonthlyCharges | +0.001 | 1.001 | $1 increase → effectively no change in churn odds (*p* = 0.72) |
| TotalCharges | ≈0 | ≈1.000 | Near-zero effect, suppressed by collinearity (*p* = 0.49) |

> **Assessment**:None of the three predictors is statistically significant, confirming that continuous financial variables alone have no linear predictive power. 
---

## 3.2 Regression in High Dimensions
### Principal Component Analysis (PCA)

```{r pca-analysis, fig.height=5}
X_pca    <- df_model[, c("Tenure", "MonthlyCharges", "TotalCharges")]
pca_result <- prcomp(X_pca, scale. = TRUE, center = TRUE)

fviz_eig(pca_result,
         addlabels = TRUE,
         barfill   = "#2196F3",
         barcolor  = "white",
         linecolor = "#F44336",
         main      = "Figure 8: PCA Scree Plot — Variance Explained by Component")
```

```{r pca-summary}
pca_summary <- summary(pca_result)$importance
kable(round(pca_summary, 4),
      caption = "Table 14: PCA variance explained for Tenure, MonthlyCharges, TotalCharges") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r pca-loadings, fig.height=5}
loadings_df <- as.data.frame(pca_result$rotation) %>%
  rownames_to_column("Variable")

kable(round(pca_result$rotation, 4),
      caption = "Table 15: PCA loadings — contribution of each variable to each component") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)

fviz_pca_biplot(pca_result,
                repel    = TRUE,
                col.var  = "#F44336",
                col.ind  = "steelblue",
                alpha.ind = 0.2,
                label    = "var",
                title    = "Figure 9: PCA Biplot — Variables and Individuals")
```

> **Interpretation**: **PC1 (63.5% variance)** captures general billing magnitude — all three variables load positively. **PC2 (34.8% variance)** contrasts MonthlyCharges (+0.82) against Tenure (−0.57).

---

## 3.3 Ridge Regression

```{r ridge-regression, fig.height=5}
X_reg <- model.matrix(
  ChurnBinary ~ Tenure + MonthlyCharges + TotalCharges + SeniorCitizen +
                Gender_enc + Partner_enc + Dependents_enc +
                PhoneService_enc + InternetService_enc + Contract_enc,
  data = df_model
)[, -1]  
X_reg_scaled <- scale(X_reg)
y_binary     <- df_model$ChurnBinary

set.seed(42)
ridge_cv <- cv.glmnet(
  x        = X_reg_scaled,
  y        = y_binary,
  alpha    = 0,
  nfolds   = 10,
  family   = "binomial",
  type.measure = "auc"
)

plot(ridge_cv,
     main = "Figure 10: Ridge Regression — Cross-Validated AUC vs. log(λ)")
```

```{r ridge-optimal}
lambda_min_ridge <- ridge_cv$lambda.min
lambda_1se_ridge <- ridge_cv$lambda.1se

cat(sprintf("Ridge λ.min:  %.4f  (AUC = %.4f)\n",
            lambda_min_ridge,
            max(ridge_cv$cvm)))
cat(sprintf("Ridge λ.1se:  %.4f\n", lambda_1se_ridge))

ridge_coefs <- as.matrix(coef(ridge_cv, s = "lambda.min"))
ridge_coef_df <- data.frame(
  Feature     = rownames(ridge_coefs),
  Coefficient = round(as.numeric(ridge_coefs), 4)
) %>%
  filter(Feature != "(Intercept)") %>%
  arrange(desc(abs(Coefficient)))

```

---

## 3.4 Lasso Regression

```{r lasso-regression, fig.height=5}
set.seed(42)
lasso_cv <- cv.glmnet(
  x            = X_reg_scaled,
  y            = y_binary,
  alpha        = 1,
  nfolds       = 10,
  family       = "binomial",
  type.measure = "auc"
)

plot(lasso_cv,
     main = "Figure 11: Lasso Regression — Cross-Validated AUC vs. log(λ)")
```

```{r lasso-coefficient-path, fig.height=5}
lasso_path <- glmnet(X_reg_scaled, y_binary, alpha = 1, family = "binomial")
plot(lasso_path, xvar = "lambda",
     main = "Figure 12: Lasso Coefficient Paths — Feature Selection as λ Increases")
```

```{r lasso-optimal}
lambda_min_lasso <- lasso_cv$lambda.min
lambda_1se_lasso <- lasso_cv$lambda.1se

cat(sprintf("Lasso λ.min:  %.4f  (AUC = %.4f)\n",
            lambda_min_lasso, max(lasso_cv$cvm)))
cat(sprintf("Lasso λ.1se:  %.4f\n", lambda_1se_lasso))

lasso_coefs <- as.matrix(coef(lasso_cv, s = "lambda.min"))
lasso_coef_df <- data.frame(
  Feature     = rownames(lasso_coefs),
  Coefficient = round(as.numeric(lasso_coefs), 4),
  Selected    = ifelse(abs(as.numeric(lasso_coefs)) > 1e-4, "Yes", "No — Zeroed")
) %>%
  filter(Feature != "(Intercept)") %>%
  arrange(desc(abs(Coefficient)))

n_zeroed <- sum(lasso_coef_df$Selected == "No — Zeroed")
cat(sprintf("\nFeatures zeroed out by Lasso: %d / %d\n", n_zeroed, nrow(lasso_coef_df)))

kable(lasso_coef_df,
      caption = paste0("Table 17: Lasso Coefficients at λ = ",
                       round(lambda_min_lasso, 4), " with feature selection status")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  column_spec(3, bold = TRUE,
              color  = ifelse(lasso_coef_df$Selected == "Yes", "#2E7D32", "#C62828")) %>%
  column_spec(2, color = ifelse(lasso_coef_df$Coefficient > 0, "#1565C0", "#C62828"))
```

### Ridge vs. Lasso Comparison

```{r ridge-lasso-compare, fig.height=4}
compare_df <- data.frame(
  Feature     = ridge_coef_df$Feature,
  Ridge       = ridge_coef_df$Coefficient,
  Lasso       = lasso_coef_df$Coefficient[match(ridge_coef_df$Feature, lasso_coef_df$Feature)]
) %>%
  pivot_longer(cols = c("Ridge", "Lasso"), names_to = "Model", values_to = "Coefficient")

ggplot(compare_df, aes(x = reorder(Feature, abs(Coefficient)), y = Coefficient,
                        fill = Model)) +
  geom_col(position = "dodge", width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = c("Ridge" = "#2196F3", "Lasso" = "#FF9800")) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  labs(title    = "Figure 13: Ridge vs. Lasso Coefficient Comparison",
       subtitle = "Standardised coefficients at optimal λ",
       x        = NULL,
       y        = "Coefficient Value")
```

```{r ridge-lasso-summary-table}
rl_compare <- data.frame(
  Property = c("Penalty Term", "Sparsity", "Feature Selection",
                "Collinearity Handling", "Optimal λ", "Zeroed Features",
                "Best Use Case"),
  Ridge = c("Σβ² (squared)", "No — shrinks but never zeros",
             "No — all features retained",
             "Groups correlated features together",
             round(lambda_min_ridge, 4), "0 / 10",
             "All features relevant; control collinearity"),
  Lasso = c("Σ|β| (absolute)", "Yes — exact zeros possible",
             "Yes — automatic variable selection",
             "Selects one from correlated group",
             round(lambda_min_lasso, 4), paste0(n_zeroed, " / 10"),
             "Sparse true model; interpretability needed")
)

kable(rl_compare, col.names = c("Property", "Ridge (L2)", "Lasso (L1)"),
      caption = "Table 18: Ridge vs. Lasso — Methodology and Results Comparison") %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  column_spec(1, bold = TRUE)
```

> **Conclusion**: At the optimal λ, both Ridge and Lasso converge to nearly identical solutions because the true predictive signal is weak and distributed across all features. The dataset presents a fundamental challenge: near-uniform churn distribution makes all linear methods struggle.

*End of Analysis*
